Rhyme: A Query Language for Nested Data Structures

### 2024-12-14 paper.tex
finish sectionn 4.2

@@

Different sparsity patterns require different ways of iterating tensors.

>>>>

Different sparsity patterns require different ways of iterating tensors, and it is non-trivial to pick the best strategy, it also adds another layer of complexity to \rhyme's loop scheduling algorithm. Originally, if a generator iterates over multiple inputs, \rhyme will generate only one loop to iterate over one of the inputs and check whether the key exists for the remaining inputs. However, in the face of sparse tensor algebra, this naive loop co-iteration algorithm may yield suboptimal performance.\par
For example, if we are performing a dot product of a sparse vector $A$ and dense vector $B$. And assume that $A$ has $m$ non-zero elements and is stored in a compressed format, and $B$ has $n$ elements. If we choose to loop over $A$ first and check existence in $B$, the computation only takes $\Theta(m)$ complexity as existence checks in dense vector takes constant time. However, if we choose to loop over $B$ first and check existence in $A$, the computation will take $\Theta(n\log{}m)$, as existence checks in sparse vector takes logarithmic time at best. This simple example shows that careless selection of co-iteration strategies in dense/sparse tensor algebra could greatly downgrade the performance.\par
The complexity continues to rise if we consider co-iteration of multiple sparse tensors. When we iterate over multiple sparse tensors, picking one sparse tensor to loop over and check existence in others will yield suboptimal performance regardless of the tensor we choose to loop over. For example, if we are performing a dot product of a sparse vector $A$ and sparse vector $B$. And assume that $A$ has $m$ non-zero elements and is stored in a compressed format, and $B$ has $n$ non-zero elements and is stored in a compressed format. And assume that the non-zero indices stored in both tensors are sorted. If we pick to loop over $A$ first and check existence in $B$, the computation will take $\Theta(m\log{}n)$ assuming we use binary search. If we pick to loop over $B$ first, the computation will take $\Theta(n\log{}m)$. However, an efficient co-iteration strategy (the two-way merge algorithm \cite{knuth}) that advancing indices in $A$ and $B$ together only need $O(m + n)$ time. This is much more efficient especially if $m$ and $n$ are at the same scale. It should be noted that we are talking about the strategy for multiplication here where all operands need to be non-zero for the result to be non-zero. Addition on the other hand will require a similar but different co-iteration strategy.\par
The N-way merge algorithm works by assigning pointers to each sparse vectors, each starting at the first non-zero indices. Then we will compare the indices in all the sparse vectors and only execute the computation when all the indices are the same, which means this index has non-zero value in all sparse vectors. If the indices are not the same, we can advance the pointer of the sparse vector that has the smallest index and that index can be skipped. We can also find the largest index among the pointers and advance all other pointers until their pointers point to indices larger or equal to the current largest index. The rationale is that all the indices smaller that the current max index must not exist in all sparse vectors, therefore can be skipped. We will apply this strategy iteratively until one of the sparse vectors runs out of value.\par
\rhyme compiler will pick an efficient strategy for different co-iteration generators depending on the dense/sparse patterns of the data sources. Combing this with the loop scheduling and fusion optimization that \rhyme's original backend offers, we can generate efficient code for sparse tensor algebra inside a multi-paradigm language framework.
Different sparsity patterns require different methods for iterating over tensors, making it non-trivial to select the best strategy. This adds an additional layer of complexity to \rhyme's loop scheduling algorithm. Originally, when a generator iterates over multiple inputs, \rhyme generates a single loop to iterate over one input and checks for the existence of corresponding keys in the remaining inputs. However, in the context of sparse tensor algebra, this naive co-iteration approach may result in suboptimal performance.\par
For example, consider the dot product of a sparse vector $A$ and a dense vector $B$. Assume that $A$ has $m$ non-zero elements stored in a compressed format, and $B$ has $n$ elements. If we loop over $A$ and check for the existence of indices in $B$, the computation takes $\Theta(m)$ time because existence checks in a dense vector occur in constant time. Conversely, if we loop over $B$ and check for the existence of indices in $A$, the computation takes $\Theta(n\log{}m)$ time because existence checks in a sparse vector require logarithmic time due to the use of binary search. This example highlights how an inappropriate co-iteration strategy can significantly degrade performance in dense/sparse tensor algebra.\par
The complexity further escalates when co-iterating over multiple sparse tensors. Simply choosing one sparse tensor to iterate over and checking for existence in others will generally yield suboptimal performance. For instance, consider the dot product of two sparse vectors $A$ and $B$. Suppose $A$ has $m$ non-zero elements and $B$ has $n$ non-zero elements, both stored in compressed formats with sorted indices. If we loop over $A$ and check existence in $B$, the computation requires $\Theta(m\log{}n)$ time using binary search. Similarly, looping over $B$ and checking in $A$ requires $\Theta(n\log{}m)$ time. However, an efficient co-iteration strategy, such as the two-way merge algorithm \cite{knuth}, advances indices in $A$ and $B$ together and achieves a time complexity of $O(m + n)$. This strategy is especially advantageous when $m$ and $n$ are of comparable scales. Note that this strategy applies to multiplication operations, where all operands must be non-zero for the result to be non-zero. For addition operations, a similar but distinct co-iteration strategy is required.\par
The N-way merge algorithm addresses co-iteration by assigning pointers to each sparse vector, starting at their respective first non-zero indices. The algorithm compares indices across all sparse vectors, executing the computation only when the all indices are the same (indicating non-zero values in all vectors at this index). If the indices do not match, the pointer of the vector with the smallest index is advanced, skipping that index. Alternatively, the algorithm can identify the largest index among the current pointers and advance all other pointers to indices greater than or equal to the largest index. This optimization skips indices that cannot exist in all sparse vectors. The process repeats iteratively until one of the sparse vectors is exhausted.\par
The \rhyme compiler selects an efficient co-iteration strategy for different generators based on the dense/sparse patterns of the data sources. Combined with \rhyme's backend loop scheduling and fusion optimizations, this approach enables the specialized generation of high-performance code for sparse tensor algebra within a multi-paradigm language framework.

@@
